---
title: "Illustration of PCAviz on POPRES data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCAviz POPRES demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we demonstrate that the PCAviz package can be used
to quickly create sophisicated plots that closely reproduce
publication-quality figures; compare the plots here against the plots
in Novembre *et al*, "Genes mirror geography within Europe," *Nature*
**456**, 274-–274 (2008). The principal components computed from
genotypes of European samples (the POPRES sample) have a remarkably
close connection to the geographic distribution of the samples, which
we examine in this vignette.

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE,comment = "#",fig.width = 7,
                      fig.height = 5,fig.align = "center",
                      fig.cap = "&nbsp;")
```

Begin by loading these packages into your R environment.

```{r, message = FALSE}
library(PCAviz)
library(magrittr)
library(cowplot)
```

## Load the data and PCA results

Load the POPRES data and PCA results, and combine these data into a
"pcaviz" object. Here we also generate a sensible abbreviation for the
country labels based on the ISO standard. (Note that abbreviated country
labels are already available in the "abbrv" column, but we generate a 
new abbreviation to illustrate one of the PCAviz package features.)

```{r}
data(popres)
popres <- pcaviz(dat = popres$x,
                 sdev = popres$sdev,
                 rotation = popres$rotation)
popres <- pcaviz_abbreviate_var(popres,"country")
```

Print a summary of the POPRES data.

```{r}
summary(popres)
```

Plot the eigenvalues of the first four PCs.

```{r, fig.height = 5,fig.width = 5}
screeplot(popres,type = "eigenvalue")
```

Examining the loadings for the first PC, a region on chromosome 15 clearly
stands out for having an unusually large loading:

```{r, fig.width = 8, fig.height = 2.5}
pcaviz_loadingsplot(popres,min.rank = 0.9,gap = 400)
```

This region is the *HERC2/OCA2* locus---a locus that has been shown in
many studies to be involved in adaptation to local enviornments through 
natural selection.

## Select data samples, and manipulate PCs

Remove the Russian (RU) and Scottish (Sct) samples, then rescale and
rotate the PCs.

```{r}
popres <- subset(popres,!(abbrv == "RU" | abbrv == "Sct")) %>%
          pcaviz_reduce_whitespace(c("PC1","PC2")) %>%
          pcaviz_rotate(105)
```

Retrieve the suggested coloring scheme for plotting the samples by
country-of-origin.

```{r}
clrs <- c(with(popres$data,tapply(as.character(color),country,
                                  function (x) x[1])))
names(clrs) <- NULL
```

## Create visualizations of the POPRES data

Calling "plot" without specifying any additional options shows the
POPRES samples projected onto the first two PCs, with the samples
labeled according to their country-of-origin. This plot closely
reproduces Fig. 1a of Novembre *et al* (2008). Note the use of the
abbreviated country names.

```{r, results = "hide", fig.width = 8}
plot(popres)
```

One difference with the published plot is the color scheme used.
In this next code chunk, we do a better job reproducing the figure in
the *Nature* paper by using the colors provided in the `popres` data
table. (Note that the Russian and Scottish samples are missing from
our plot since we removed them above.)

```{r, results = "hide", fig.width = 8}
plot(popres,colors = clrs)
```

Compare this projection onto PCs 1 and 2 against the countries'
actual geographical locations:

```{r, fig.width = 5, fig.height = 5}
plot(popres,coords = c("longitude","latitude"),group = "country.abbrv",
     label = NULL,colors = clrs,overlay = overlay_map_europe,
     show.legend = FALSE)
```

Next, we examine the relationships between PC1 and longitude, and
between PC2 and latitude. When plotting a PC against a data column,
the linear best fit is automatically drawn (dashed line), along
with confidence intervals (dotted lines).

```{r, results = "hide", fig.height = 3.5}
plot1 <- plot(popres,coords = c("latitude","PC2"),group = NULL,
              show.legend = FALSE)
plot2 <- plot(popres,coords = c("longitude","PC1"),group = NULL,
			        show.legend = FALSE)
plot_grid(plot1,plot2,labels = c("A","B"))
```

This is a more condensed version of the previous plot obtained by
setting `group = "country.abbrv"` and `label = NULL`.

```{r, fig.height = 3.5}
plot1 <- plot(popres,coords = c("latitude","PC2"),group = "country.abbrv",
              label = NULL,show.legend = FALSE)
plot2 <- plot(popres,coords = c("longitude","PC1"),group = "country.abbrv",
              label = NULL,show.legend = FALSE)
plot_grid(plot1,plot2,labels = c("A","B"))
```

Next, we set the "color" argument to visualize the East-West and 
North-South trends in PC space.

```{r, results = "hide", fig.width = 8, fig.height = 4}
plot1 <- plot(popres,color = "latitude",group = "country.abbrv")
plot2 <- plot(popres,color = "longitude",group = "country.abbrv")
plot_grid(plot1,plot2,labels = c("A","B"))
```

Here is alternative way to visualize the same result on a map of
Europe; the colors are used to depict PCs 1 and 2.

```{r, results = "hide", fig.width = 8, fig.height = 3.5}
plot1 <- plot(popres,coords = c("longitude","latitude"),color = "PC1",
              group = "country.abbrv",overlay = overlay_map_europe)
plot2 <- plot(popres,coords = c("longitude","latitude"),color = "PC2",
              group = "country.abbrv",overlay = overlay_map_europe)
plot_grid(plot1,plot2,labels = c("A","B"))
```

In view of the strong relationship between geography within Europe and
the first two PCs computed from the genotype data, we attempt to fit the
samples onto a map of Europe by translating and scaling PCs 1 and 2.

```{r}
fit1 <- lm(longitude ~ PC1,popres$data)
fit2 <- lm(latitude ~ PC2,popres$data)
mu1  <- coef(fit1)[["(Intercept)"]]
mu2  <- coef(fit2)[["(Intercept)"]]
b1   <- coef(fit1)[["PC1"]]
b2   <- coef(fit2)[["PC2"]]
popres.for.map <- 
  pcaviz_scale(popres, scale = c(b1,b1),dims = c("PC1","PC2")) %>%
  pcaviz_translate(a = c(mu1,mu2),dims = c("PC1","PC2"))
```

Plot PCs 1 and 2 onto the map of Europe after this scaling and 
translation step.

```{r, fig.width = 6}
plot(popres.for.map,label = NULL,group = "country",
     show.legend = FALSE,overlay = overlay_map_europe)
```

Next, plot all the samples onto the map of Europe.

```{r, results = "hide", fig.width = 9}
plot(popres.for.map,group = NULL,overlay = overlay_map_europe)
```

We can also easily create interactive version of this plot, and embed it 
in a separate HTML file. View the interactive plot [here](popres_plotly.html).

```{r, results = "hide"}
popres_plotly <- 
  plot(popres.for.map,plotly = TRUE,overlay = overlay_map_europe,
       tooltip = c("id","country","longitude","latitude","PC1","PC2"),
       plotly.file = "popres_plotly.html")
```

Note that the interactive plot can also be easily embedded within this 
document. In this example we have placed it in a separate webpage because
loading the JavaScript can be slow in some browsers.

Finally, we compare PCs 1 and 2 against country-of-origin using the
"pcaviz_violin" interface. Since density estimates are less useful for
small numbers of samples, here we show only countries with a sample
size of at least 10.

```{r, fig.width = 6.5, fig.height = 7}
countries <- c("AT","BE","CH","CZ","DE","ES","FR","GB","HU","IE","IT",
               "NL","PL","PT","RO","RS","SE")
dat <- subset(popres,is.element(abbrv,countries))
pcaviz_violin(dat,pc.dims = c("PC1","PC2"),
              plot.grid.params = list(nrow = 2))
```
