---
title: "Illustration of PCAviz on RegMap data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCAviz RegMap demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates PCAviz for visualizing principal components
in the Regmap data set. The first two PCs of the genotype data
distinguish geographic origins of the *Arabidopsis thaliana* samples,
although some overlap is evident.

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = TRUE,comment = "#",fig.width = 6,
                      fig.height = 5,fig.align = "center",
                      fig.cap = "&nbsp;",results = "hold")
```

Begin by loading these packages into your R environment.

```{r, message = FALSE}
library(PCAviz)
library(cowplot)
```

## Load the data and PCA results

Load the RegMap data and the computed PCs, and convert these data to a
"pcaviz" object. (We do not include the "region" column.)

```{r}
data(regmap)
regmap <- pcaviz(dat = regmap[-10])
```

Print a summary of the RegMap data.

```{r}
summary(regmap)
```

## Create visualizations of the RegMap PCA results

Calling "plot" without specifying any additional options shows the
projection of the samples onto the first two PCs, with the samples
labeled by the country in which they were found.

```{r}
plot(regmap)
```

The group summaries add a lot of clutter to these plots, so
we remove them. We also don't need the legend.

```{r}
plot(regmap,group = NULL,show.legend = FALSE)
```

Next, show the country assignments using different colors and shapes 
instead of labels.

```{r, fig.width = 7, fig.height = 6}
plot(regmap,draw.points = TRUE,group = NULL)
```

By setting the `color` argument to "longitude" and "latitude", these
next two plots help us understand the relationship between geography
and projection onto the PC embedding.

```{r, fig.width = 7, fig.height = 3}
plot1 <- plot(regmap,draw.points = TRUE,color = "longitude",group = NULL)
plot2 <- plot(regmap,draw.points = TRUE,color = "latitude",group = NULL)
plot_grid(plot1,plot2,labels = c("A","B"))
```

Does "median intensity" (this is a technical artifact) explain some
variation in the first PC?

```{r, fig.width = 7, fig.height = 6}
plot(regmap,coords = c("PC1","median_intensity"),draw.points = TRUE,
     group = NULL)
```

The `plot` function can also be used to quickly plot combinations of PCs. 
This code also illustrates customization of the plotting parameters---in this 
case, since the plots are small, the default point size is a bit too large.

```{r, fig.width = 7, fig.height = 7}
plot(regmap,coords = paste0("PC",1:4),group = NULL,draw.points = TRUE,
     geom.point.params = list(size = 1,na.rm = TRUE))
```
